---
title: "M08-Reflection Essay-Advanced Data Types"
subtitle: "Revealjs Presentation"
author: "Alejandro Camara"
institute: "IBM 3202.01, Cal Poly Pomona"
date: today
date-format: "M/D/YYYY"
format: 
  revealjs:
    theme: serif
    width: 1600
    height: 900
    footer: Title of Your Project
    transition: slide
    transition-speed: default
    incremental: false
    toc: true
    toc-depth: 1
    slide-number: true
    scrollable: true
    smaller: true
    code-fold: false
    code-overflow: wrap
    number-sections: false
    number-depth: 5
    embed-resources: true
    css: styles.css
editor: visual
execute: 
  echo: true
  freeze: auto
---

# 1. Pick three functions ***Dr. Dave Robinson*** used

> In Step 1, ***Dr. Dave Robinson*** demonstrated how you can deal with various data types using string, forecast, and lubricate packages. Pick three functions he used and how it helped him achieve his wrangling goals.

## I chose the following functions:

::: incremental
1.  mutate(column = ifelse(`str_detect`(metric, "standard_errors\_"), "standard_error", "percent),<br>metric = `str_remove(metric, "standard_errors_"))`: He wanted to reshape the data so there is one row for each year and each race. Overall, using `str_detect` allowed **Dr. Robinson** to achieve his wrangling goals as he was able to locate and later remove the standard errors column, allowing him to now spread the value across the column.

2.  filter(year \>= 1940,<br>`!str_detect(race_ethnicity, "Islander -"))`: Here, **Dr. Robinson** noticed that, by looking at the legend, three variables **Asian/Pacific Islander - Asian, Total - Asian/Pacific Islander** and **Asian/Pacific Islander - Pacific Islander** were shown separately on the line chart. To clean this up, he decided to go only with **Total - Asian/Pacific Islander** since this had more data than the other two together. Overall, using `!str_detect` allowed **Dr. Robinson** to achieve his wrangling goals as he was able to remove the unwanted variables.

3.  mutate(degree = fct_relevel(degree, "High School"),<br>race_ethnicity = `str_remove(race_ethnicity, "Total - "))`: Subsequently, he removed the word **Total -** because he believed that it was confusing. Instead of **Total - Asian/Pacific Islander** he preferred for it to show **Asian/Pacific Islander.** Overall, using `str_remove` allowed **Dr. Robinson** to achieve his wrangling goals as the variable name was edited to remove any confusion for the audience.
:::

# 2. Summarize the presentation by ***Dr. Pat Scholoss***

> Summarize the presentation by ***Dr. Pat Scholoss***, a Professor at the University of Michigan, on how to deal with date data in Step 2. How much were you able to understand his presentation? compare his style with Dave Robinson's. What do you like or dislike?

## What I understood from the presentation:

To start off, ***Dr. Pat Scholoss,*** mentions that his son has a business where he bales hay. So, he needs it to be dry to bale and wet for the hay to grow. His question is "As of today, July 22nd, has the year been wetter than it has been based on other years that we have data recorded for." Ultimately, he will be using group by and summarize and also use date related functions to create an attractive visual for his son, demonstrating whether or not its been an abnormally wet or dry year. Next up, he shows his code `source("code/local_weather.R")` which is a r script that will download data from his local weather station, which is Ann Arbor, Michigan. His data goes back to 1891. ***Dr. Scholoss*** now loads the r script, which also includes library call for the tidy verse, lubridate as well as the glue package. He then gets his local weather data frame, `local_weather %>%`, which shows the following columns: date, tmax, prcp, and snow. He is only interested in the amount of precipitation and does a select function on that, `select(date, prcp) %>%`. He also wants to know the year, month and the day, so he adds a mutate function,<br>`mutate(year = year(date),`<br>`month = month(date),`<br> `day = day(date))`, the year function comes from lubridate. Since there are NA values for the precipitation column, he decided to drop them by using `drop_na(prcp) %>%`, adding this below select. Since the 1891 data starts in October, he also drops that by using `filter(year != 1891)`, resulting in his data now starting on January 1st, 1892. He wants a plot that has the different months on the x-axis and cumulative precipitation total on the y-axis. Each line representing each year in gray and 2022 in blue to see how it compares to the rest. To make things easier, he decides to remove February 29th. To do so, he adds to his filter function, `filter(year != 1891) & !(month == 2 & day ==29)) %>%`. After removing leap year days, he reiterates that he wants cumulative precipitation total by year, so he uses `group_by(year)`. As a result, we can now see 131 year groups, 130 and the year 2022. He then does `mutate(cum_prcp = cumsum(prcp)) %>%`, which gives us the date and cumsum. After that, he ungroups his data, `ungroup()`. Now he wants to feed it to ggplot, cum_prcp for y-axis, grouping the data by the year, and 1 year worth of date for x-axis, since he has 131 years worth of dates. So for the x-axis, he decides to turn the year column into 2022 and add the month and day, ultimately creating a pseudo date. To do so, use `mutate(new_date = ymd(glue("2022-{month}-{day}"))) %>%`. This adds a new_date column. ymd is another function from lubridate that will take dates formatted like this into a "Year-Month-Day" format. Instead of being type \<glue\> its now type \<date\>. Now he feeds this into ggplot: `ggplot(aes(x = new_date, y = cum_prcp, group = year)) +`<br>`geom_line()`. At the very top before local_weather, he creates a variable, `this_year <- today())`. He also adds `is_this_year = year == this_year` to mutate. He decides to color the lines by is this year, adding `color = is_this_year` to ggplot function. By doing this, we can now see the year 2022 in blue and the other 130 years in red. ***Dr. Scholoss*** interprets what he sees in the visualization that it seems a little bit wetter than average. Moreover, he dicededs that he wants to make the plot more attractive. He begins by adding `scale_color_manual(breaks = c(F, T),`<br>`values c("lightgray", "dodgerblue"))`. To make the year 2022 thicker than the others, he adds `size=is_this_year` in the ggplot function. He recieved a warning message: *Using size for a discrete variable is not advised*. So to follow what the message advised, he adds `scale_size_manual(breaks = c(F, T)`<br>`values = c(0.3, 1))`. To remove the legend he adds `show.legend = False` to `geom_line()`. Next, he saves the plot to a file, `ggsave("figures/cumulative_prcp.png", width = 6, height = 5)`. Next, he focuses on cleaning up the x-axis. He mentions that ggplot recognizes that the new date column was of type date and treated that as dates, January, April, July, October, January of 2022 to 2023. He just wants to have the month and does so by using this function: `scale_x_date(date_labels = %B, date_breaks = "2 months")`. This shows the x-axis only with the full name of the month. date_breaks shows the dates by every 2 months: February, April, June, August, October, and December. With scale x date, we are able to manipulate the appearance of the dates on the x-axis as well as on the y-axis if needed. Upon that, the focus shifts on cleaning up the theming of the figure. He decided to turn off the grid lines as well as the background and adds x and y axis. He does so by, adding this function below scale x date, `theme(panel.background = element_blank(), axis.line = element_line(), panel.grid = element_blank.` In regards to formatting y-axis, he mentions that he wants to remove the gap at the bottom of the 0 since we would never get a negative cumulative precipitation. Below scale x date, he includes this function: `scale_y_continuous(breaks = seq(0, 1200, 300), labels = seq(0, 120, 30), limits = c(0, 1200), expand = c(0, 0))`. This turns off the expansion. He now transforms the y scale into centimeters by using labels and breaks. Right below he adds `labs(x = NULL, y = "Cumulative Precipitation (cm)")`. To determine if 2022 is above or below what we would expect at this point in the year he added a line through the data to indicate whether or not that's the average. He uses `geom_smooth(aes(group = 1), color = "black", size = 0.3, se = False)`. Finally, he talks about adding and formatting a title. So he goes back to labs and adds `title ="Through July 22nd, the cumulative precipitation near Ann Arbor, MI is <span style = 'color: dodgerblue'>above average</span> for 2022"`. So, the title goes off the right side of the screen. He wants the title to be left justified on the plot. In the theme he adds `plot.title.position = "plot",`<br>`plot.title = element_textbox_simple(margin=margin(b=10))`. I was able to understand almost everything from his presentation because of the way he explains each detail. When comparing his style with Dave Robinson's, I like how he explains what he is doing. While Dave is knowledgeable, I feel like he doesn't explain each thing he does very well.

# 3. Summarize the gtsummary package capabilities

> In Step 3, I introduced the common usage of gtsummary, which can be useful for school work. Summarize the package's capabilities as explained. How may the {gtsummary} benefit you for your school work or career?

## The gtsummary package

In the lecture, Dr. Jung uses the the trial data frame with 200 rows, one row per patient that is included in the `gtSummary` package. The variables are

- **trt**: Chemotherapy Treatment

- **age**: Age

- **marker**: Marker level (ng/mL)

- **stage**: **T Stage**: in cancer describes the size and extent of the primary tumor.

- **grade**: describes how abnormal the cancer cells look under a microscope compared to normal cells. The higher the grade, indicates more rapid growth and a greater likelihood of spread.

- **response**: Tumor Response

- **death**: Patient Died

- **ttdeath**: Months to Death/Censor

**Capabilities**

So, in showing the package's capabilities while utilizing this data, we can use `tbl_summary()`, which is designed for descriptive statistics. Here, we can look at the data, `trial`. Then,<br>`trial2 <- trial |>`<br>`select(trt, age, grade, response)`only choose specific columns from the dataset. Lastly,<br>`trial2 |>`<br>`tbl_summary()` essentially takes trial2 and summarizes each column, producing a publishable table. Another **capability** islooking at **"Differences between groups."** There are multiple ways to do so, such as Wilcoxon Rank sum test, which uses median, Welch Two Samples T-test, Chi-square with `tbl_summary`, and Chi-square with `tbl_cross()`, which is cross-tabulation with trial data. The {gtsummary} may benefit me for my school work or career in many ways. For both school and work, when working with data, we must eventually present it to others. A nicely laid out table showing the summary of the data gives the audience a quick recap of the content.

# 4. Extended presentation of the {gtsummary} package by ***Dr. Daniel Dsjoberg***

> In Step 4, you watched the extended presentation of the {gtsummary} package by ***Dr. Daniel Dsjoberg***, the very creator of the package. How does it differ from gt and gtExtras? Give three things you learned newly that were not explained in the lecture in Step 3.

## Three things I learned

::: incremental
While the three packages are related, they serve different roles in building a table. **They differ in the following ways:**

- **`gt`**: gt package created by RStudio which is for creating and styling tables from scratch.

- **`gtsummary`**: gtsummary package uses gt in the background to quickly create beautiful statistical summary tables.

- **`gtExtras`**: gtExtras package allows you to add additional visual features to gt tables. You are able to add Inline bar charts, Sparklines, and Icons.

**Three new things that I learned:**

1.  Create tabular summaries with sensible defaults but highly customizable.

2.  When writing reports, we are reporting numbers from the tables, so there is an inline functionality where you can pass an entire table object and tell gt summary "here is a table, I need to pull out the odds ratio for age and this confidence interval." Doing so will put it right into the r markdown report.

3.  It helps to label variables in the dataset since they are carried through when using functions like `tbl_summary()`.
:::

# 5. Pick one new interesting function from each of the three topics -- strings, factor, and date-time

> In Step 5, you explored my advanced codebook. Pick one new interesting function from each of the three topics -- strings, factor, and date-time --  that was not introduced in M07. Describe what they are and what they can do for you. While doing so, use actual codes and outputs. 

## New functions learned

[**Strings (`stringr`)**]{.underline}

- `str_sub()`: This function is used to **extract or modify substrings** (parts of strings) based on their positions. It’s really useful when you want consistent, readable string manipulation.

<details>

<summary>Click to reveal code and output</summary>

```{r}
library(tidyverse)

x <- c("California", "Arizona", "New York")
str_sub(x, 1, 3) # argument: start, end

# negative numbers count backwards from end
str_sub(x, -3, -1)

```

</details>

[**Factor (`forcats`)**]{.underline}

- `parse_factor()`: This function is used to convert a character vector into a factor with predefined levels. It is useful because it lets you control both the allowed values and their order when working with categorical data.

<details>

<summary>Click to reveal code and output</summary>

```{r}

x <- c("Dec", "May", "Sep")
readr::parse_factor(x, levels = month.abb)

```

</details>

[**Date-Time (`lubridate`)**]{.underline}

- `round_date()`: This function is used to round date or date-time values to the nearest specified unit of time, such as the nearest hour, day, month, or year. It is useful for simplifying and standardizing date-time data, when you want to group or analyze data at a certain time level.

<details>

<summary>Click to reveal code and output</summary>

```{r}

x <- ymd_hms("2023-03-15 10:23:45")

round_date(x, "hour")
```

</details>

# 6. What did you like the most about working with the tools you learned in this module?

> I will repeat this again here. Data wrangling is like laying a foundation/basement for a building, while visualization is building a structure above the basement. working in a basement is not a glamorous job at all. No one notices it; people will see what they can see above the ground and judge it, but without the hard work you did in the basement, the building will never stand well. Wrangling with strings and date-time data is particularly hard because we don't get to use it as often as we use numeric data. However, you will get to use factors often, so get used to them soon. What did you like the most about working with the tools you learned in this module? Elaborate on your point.

## The tools from the module

What I liked the most about working with the tools that I learned in this module was the amazing things that can be done. For instance, I liked that you can do lots of things with the string package, such as removing unwanted text in the plots legend (`str_remove`), as shown in Step 1 by ***Dr. Dave Robinson.*** Also, I liked how the **`gtsummary`** allows you to create a statistical summary table, saving you time from manually doing the calculations.

# 7. How much time have you been able to study for this module?

> By this time, I hope you found a way to deal with your challenges well. If you still have a problem and want to consult with me, you can come to see me during my office hours. How much time have you been able to study for this module? Which steps have you been able to study thoroughly?

## Time spent studying

I have been able to study a great amount of time for this module. I studied most steps thoroughly, with the exception of the advanced code book in step 5, as it was complex and had so many functions.